---
title: "Pharmacokinetics Scroll Story"
subtitle: "Wonderful Wednesdays • 2026-06-10 challenge"
format:
  html:
    toc: true
    toc-depth: 2
    smooth-scroll: true
    code-fold: true
    page-layout: full
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(plotly)
library(htmltools)

data_file <- file.path("..", "..", "..", "data", "2026", "2026-06-10", "Simulated_PK_Data.xlsx")
if (!file.exists(data_file)) {
  stop(paste0("Data file not found: ", normalizePath(data_file, winslash = "/", mustWork = FALSE)))
}

pk_data <- readxl::read_excel(data_file) |>
  dplyr::rename_with(tolower) |>
  dplyr::mutate(
    subject = as.character(subject),
    time_hr = as.numeric(time_hr),
    concentration_ng_ml = as.numeric(concentration_ng_ml)
  ) |>
  dplyr::filter(!is.na(subject), !is.na(time_hr), !is.na(concentration_ng_ml), concentration_ng_ml > 0) |>
  dplyr::arrange(subject, time_hr)

summary_by_time <- pk_data |>
  dplyr::group_by(time_hr) |>
  dplyr::summarise(
    q25 = quantile(concentration_ng_ml, 0.25),
    median = median(concentration_ng_ml),
    q75 = quantile(concentration_ng_ml, 0.75),
    p10 = quantile(concentration_ng_ml, 0.10),
    p90 = quantile(concentration_ng_ml, 0.90),
    .groups = "drop"
  )

normalized_profiles <- pk_data |>
  dplyr::group_by(subject) |>
  dplyr::mutate(
    cmax = max(concentration_ng_ml),
    concentration_norm = concentration_ng_ml / cmax
  ) |>
  dplyr::ungroup()

normalized_summary <- normalized_profiles |>
  dplyr::group_by(time_hr) |>
  dplyr::summarise(
    median_norm = median(concentration_norm),
    .groups = "drop"
  )

# --- Per-subject peak behaviour (Cmax / Tmax) ---
subject_peaks <- pk_data |>
  dplyr::group_by(subject) |>
  dplyr::arrange(time_hr, .by_group = TRUE) |>
  dplyr::filter(concentration_ng_ml == max(concentration_ng_ml)) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::rename(cmax = concentration_ng_ml, tmax = time_hr) |>
  dplyr::select(subject, tmax, cmax)

# --- Per-subject exposure: trapezoidal AUC over the observed window ---
auc_by_subject <- pk_data |>
  dplyr::group_by(subject) |>
  dplyr::arrange(time_hr, .by_group = TRUE) |>
  dplyr::summarise(
    auc = sum(diff(time_hr) * (head(concentration_ng_ml, -1) + tail(concentration_ng_ml, -1)) / 2),
    .groups = "drop"
  )
t_first <- min(pk_data$time_hr)
t_last <- max(pk_data$time_hr)

# --- Apparent terminal half-life from the log-linear tail (time >= 8 h) ---
est_lambda <- function(t, conc) {
  ok <- is.finite(conc) & conc > 0
  if (sum(ok) < 2) return(NA_real_)
  slope <- unname(coef(stats::lm(log(conc[ok]) ~ t[ok]))[2])
  if (!is.finite(slope) || slope >= 0) return(NA_real_)
  -slope
}
half_life_by_subject <- pk_data |>
  dplyr::filter(time_hr >= 8) |>
  dplyr::group_by(subject) |>
  dplyr::summarise(lambda_z = est_lambda(time_hr, concentration_ng_ml), .groups = "drop") |>
  dplyr::mutate(t_half = log(2) / lambda_z)

# --- Scalars for callouts ---
n_subject <- dplyr::n_distinct(pk_data$subject)
n_time <- dplyr::n_distinct(pk_data$time_hr)
time_points <- paste(sort(unique(pk_data$time_hr)), collapse = ", ")
t_peak <- summary_by_time$time_hr[which.max(summary_by_time$median)]
c_peak <- max(summary_by_time$median)
y_floor <- min(pk_data$concentration_ng_ml, na.rm = TRUE)
y_cap <- max(pk_data$concentration_ng_ml, na.rm = TRUE) * 1.08
conc_span <- max(pk_data$concentration_ng_ml) / min(pk_data$concentration_ng_ml)

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d)
# One font family across every chart, matching the story text (CSS body font).
story_font <- "Iowan Old Style, Palatino Linotype, Palatino, Book Antiqua, Georgia, serif"
cmax_med <- median(subject_peaks$cmax)
cmax_q25 <- quantile(subject_peaks$cmax, 0.25)
cmax_q75 <- quantile(subject_peaks$cmax, 0.75)
tmax_med <- median(subject_peaks$tmax)
tmax_tab <- sort(table(subject_peaks$tmax), decreasing = TRUE)
tmax_mode <- as.numeric(names(tmax_tab)[1])
tmax_mode_pct <- round(100 * tmax_tab[1] / n_subject)
auc_med <- median(auc_by_subject$auc)
auc_q25 <- quantile(auc_by_subject$auc, 0.25)
auc_q75 <- quantile(auc_by_subject$auc, 0.75)
thalf_med <- median(half_life_by_subject$t_half, na.rm = TRUE)
thalf_n <- sum(is.finite(half_life_by_subject$t_half))

# JS: on hover, highlight the hovered line (orange, thicker) and fade all others.
# Restores original colours/widths on unhover. Attached via htmlwidgets::onRender.
highlight_js <- "
function(el) {
  var n = el.data.length;
  var origCol = el.data.map(function(t){ return (t.line && t.line.color) || null; });
  var origW = el.data.map(function(t){ return (t.line && t.line.width) || 1; });
  el.on('plotly_hover', function(d) {
    var cn = d.points[0].curveNumber;
    var op = [], col = [], w = [];
    for (var i = 0; i < n; i++) {
      if (i === cn) { op.push(1); col.push('#D55E00'); w.push(3); }
      else { op.push(0.08); col.push(origCol[i]); w.push(origW[i]); }
    }
    Plotly.restyle(el, {'opacity': op, 'line.color': col, 'line.width': w});
  });
  el.on('plotly_unhover', function() {
    var op = []; for (var i = 0; i < n; i++) { op.push(1); }
    Plotly.restyle(el, {'opacity': op, 'line.color': origCol, 'line.width': origW});
  });
}
"

# JS: force each plot to (re)fit its container width — kills the fixed-width
# SVG that overflows into a horizontal-scroll box. Observes the container so it
# fits on first paint and on any resize (window OR column width change).
fit_js <- "
function(el) {
  function fit() { if (window.Plotly) { Plotly.Plots.resize(el); } }
  fit();
  setTimeout(fit, 0);
  setTimeout(fit, 200);
  if (window.ResizeObserver) { new ResizeObserver(fit).observe(el.parentElement); }
  window.addEventListener('resize', fit);
}
"
```

```{=html}
<style>
/* --- Full-viewport scroll-snap story --- */
html, body {
  scroll-snap-type: y proximity;
  scroll-behavior: smooth;
  scroll-padding-top: 0;
  overflow-x: hidden;                 /* no accidental left/right scroll */
}
/* Storytelling typography — elegant web-safe serif, sans-serif headings */
body, .snap-panel {
  font-family: "Iowan Old Style", "Palatino Linotype", Palatino, "Book Antiqua", Georgia, serif;
}
h1, h2, h3, .title, .subtitle, .scroll-cue {
  font-family: "Iowan Old Style", "Palatino Linotype", Palatino, "Book Antiqua", Georgia, serif;
}
.snap-panel {
  min-height: 94vh;
  scroll-snap-align: center;
  display: flex;
  flex-direction: column;
  justify-content: center;
  padding: 0.75rem 0;
  box-sizing: border-box;
  font-size: 1.2rem;
  line-height: 1.6;
  max-width: 1280px;          /* wide canvas for charts */
  margin-left: auto;
  margin-right: auto;
}
/* min-width:0 lets flex children actually shrink to the column width, so the
   chart never spills into a horizontal-scroll box */
.snap-panel > *, .snap-panel .cell, .snap-panel .cell-output-display {
  width: 100%; max-width: 100%; min-width: 0;
}
.snap-panel .cell-output-display { overflow-x: hidden; }
/* narrative text stays at a readable measure; charts go full panel width */
.snap-panel > h2, .snap-panel > p, .snap-panel > ul, .snap-panel > ol,
.snap-panel > blockquote, .snap-panel > table, .snap-panel > .scroll-cue {
  max-width: 820px; margin-left: auto; margin-right: auto;
}
.snap-panel h2 { margin-top: 0; font-size: 2rem; }
.snap-panel p, .snap-panel li, .snap-panel td, .snap-panel th { font-size: 1.18rem; }
/* charts fill the full (wide) panel width and refit — never a scroll box */
.snap-panel .cell, .snap-panel .cell-output-display,
.snap-panel .html-widget, .snap-panel .plotly {
  width: 100% !important; max-width: 100% !important;
}
/* scroll cue */
.scroll-cue { text-align: center; color: #8a8a8a; font-size: 1rem; margin-top: 0.8rem; }
.scroll-cue::after { content: " \2193"; }

/* --- Floating right-hand table of contents --- */
#TOC {
  position: fixed !important;
  top: 1.5rem;
  right: 1rem;
  width: 235px;
  max-height: 90vh;
  overflow-y: auto;
  margin: 0;
  padding: 0.6rem 0.95rem;
  background: rgba(255, 255, 255, 0.92);
  border-left: 3px solid #d8e3ef;
  border-radius: 4px;
  z-index: 1050;
  font-size: 0.92rem;
}
#TOC #toc-title { font-size: 1rem; margin: 0 0 0.4rem; font-weight: 600; }
/* Bootstrap's .collapse hides the list by default — force it open here */
#TOC ul.collapse, #TOC ul { display: block !important; }
#TOC ul { list-style: none; padding-left: 0.4rem; margin: 0; }
#TOC ul ul { padding-left: 0.8rem; }
#TOC li { margin: 0.18rem 0; line-height: 1.3; }
#TOC .nav-link { text-decoration: none; color: #555; }
#TOC .nav-link.active { color: #003f88; font-weight: 600; }
/* reserve a right gutter so charts never sit under the floating TOC */
@media (min-width: 1200px) { body { padding-right: 275px; } }
@media (max-width: 1199px) { #TOC { display: none !important; } }
</style>
```

::: {.snap-panel}
## Why this story {#why-story}

A single oral dose, `r n_subject` subjects, plasma concentration followed for 24 hours. Scroll through four questions, one full screen at a time:

1. What is the **population-level PK pattern** over time?
2. How much **between-subject variability** is present at each time point?
3. How consistent are **peak timing (Tmax)** and **peak concentration (Cmax)** across subjects?
4. Is the overall **profile shape** consistent after normalizing by subject peak?

Dataset scope: `r n_subject` subjects, `r n_time` post-dose time points (`r time_points` h) up to 24 hours.

::: {.scroll-cue}
scroll to begin
:::
:::

::: {.snap-panel}
## At a glance {#at-a-glance}

Population-level PK metrics (per-subject values summarized as median [IQR]):

| Metric | Value |
|---|---|
| Subjects × time points | `r n_subject` × `r n_time` (single oral dose, 0–24 h) |
| Median Cmax | `r fmt(cmax_med, 1)` ng/mL [`r fmt(cmax_q25, 1)`–`r fmt(cmax_q75, 1)`] |
| Median Tmax | `r fmt(tmax_med, 1)` h (`r tmax_mode_pct`% of subjects peak at `r fmt(tmax_mode, 0)` h) |
| Median AUC(`r fmt(t_first,1)`–`r fmt(t_last,0)` h) | `r fmt(auc_med, 0)` ng·h/mL [`r fmt(auc_q25, 0)`–`r fmt(auc_q75, 0)`] |
| Apparent terminal t½ (from ≥8 h) | `r fmt(thalf_med, 1)` h (n = `r thalf_n`) |

> AUC is trapezoidal over the observed sampling window (first sample at `r fmt(t_first,1)` h, so this is not extrapolated to time 0 or infinity). Half-life is an *apparent* terminal estimate from the log-linear tail and is provided as an orientation value only.
:::

::: {.snap-panel}
## Individual trajectories {#population-signal}

**Hover any line** to highlight that subject and read their ID and values — all other lines fade out. The visual noise is the message: `r n_subject` subjects, all different.

```{r}
# Build spaghetti using per-subject loop — avoids formula lazyeval crash
subjects <- sort(unique(pk_data$subject))

p_spaghetti <- plotly::plot_ly()
for (s in subjects) {
  d <- pk_data[pk_data$subject == s, ]
  ht <- paste0("<b>Subject ", s, "</b><br>Time: %{x} h<br>Concentration: %{y:.2f} ng/mL<extra></extra>")
  p_spaghetti <- plotly::add_trace(
    p_spaghetti,
    x = d$time_hr, y = d$concentration_ng_ml,
    name = s, type = "scatter", mode = "lines+markers",
    showlegend = FALSE,
    line = list(color = "rgba(80,80,80,0.45)", width = 1),
    marker = list(size = 10, opacity = 0),
    hovertemplate = ht
  )
}

p_spaghetti <- p_spaghetti |>
  plotly::layout(
    title = list(text = "Individual trajectories — hover to identify a subject (n = 50)",
                 font = list(size = 15), x = 0, xanchor = "left"),
    xaxis = list(title = "Time post-dose (hours)", automargin = TRUE),
    yaxis = list(title = "Concentration (ng/mL)", range = c(0, y_cap), automargin = TRUE),
    hovermode = "closest",
    margin = list(t = 55, r = 20, b = 55, l = 75),
    showlegend = FALSE,
    font = list(family = story_font, size = 14),
    height = 470
  )

p_spaghetti |> plotly::config(responsive = TRUE) |>
  htmlwidgets::onRender(fit_js) |> htmlwidgets::onRender(highlight_js)
```
:::

::: {.snap-panel}
## Population fan — linear axis {#fan-linear}

The same variability distilled into a clean quantile fan. **The fan width at each time is the key signal.** On the linear axis the rise-to-peak reads directly.

```{r}
# Summary fan — built with direct vectors, no formula evaluation.
# Rendered on both linear and log y-axes as two panels: a build-time log axis
# renders reliably, whereas a relayout log toggle did not.
build_fan <- function(log_scale = FALSE, show_legend = TRUE) {
  if (log_scale) {
    # Conventional 1-2-5 log breaks bracketing the observed 2.8-85.8 ng/mL range.
    # plotly's default log labelling prints each decade's mantissa (…8,9,10,then 2,3…),
    # which reads as a confusing "3-10 then 2-9"; explicit breaks give a clean axis.
    log_ticks <- c(2, 5, 10, 20, 50, 100)
    # plotly log-axis quirk: `range` is in log10 units, but `tickvals` are RAW
    # data values (plotly logs them internally). Passing log10(ticks) here places
    # ticks off-range and hides all labels.
    y_axis <- list(title = "Concentration (ng/mL, log scale)", type = "log",
                   range = c(log10(2), log10(100)),
                   tickmode = "array", tickvals = log_ticks,
                   ticktext = as.character(log_ticks), automargin = TRUE)
    ttl <- "Population summary on a log y-axis — reads the elimination phase (n = 50)"
  } else {
    y_axis <- list(title = "Concentration (ng/mL)", range = c(0, y_cap), automargin = TRUE)
    ttl <- "Population summary: quantile fan (P10–P90, IQR) + median (n = 50)"
  }
  # Fully pre-formatted hover string per time point — attached to the visible
  # median markers with hoverinfo="text" and hovermode="closest". No customdata
  # indexing and no x-unified snapping, so each point shows exactly its own stats.
  hover_txt <- sprintf(
    "Time %g h<br>Median: %.2f ng/mL<br>IQR: %.2f–%.2f<br>P10–P90: %.2f–%.2f",
    summary_by_time$time_hr, summary_by_time$median,
    summary_by_time$q25, summary_by_time$q75,
    summary_by_time$p10, summary_by_time$p90
  )
  plotly::plot_ly() |>
    plotly::add_ribbons(
      x = summary_by_time$time_hr, ymin = summary_by_time$p10, ymax = summary_by_time$p90,
      name = "P10–P90", fillcolor = "rgba(191,219,247,0.40)", showlegend = show_legend,
      line = list(color = "rgba(191,219,247,0)"), hoverinfo = "skip"
    ) |>
    plotly::add_ribbons(
      x = summary_by_time$time_hr, ymin = summary_by_time$q25, ymax = summary_by_time$q75,
      name = "IQR (Q1–Q3)", fillcolor = "rgba(86,180,233,0.55)", showlegend = show_legend,
      line = list(color = "rgba(86,180,233,0)"), hoverinfo = "skip"
    ) |>
    plotly::add_trace(
      x = summary_by_time$time_hr, y = summary_by_time$median,
      type = "scatter", mode = "lines+markers",
      name = "Median", line = list(color = "#003f88", width = 2.5),
      marker = list(color = "#003f88", size = 7), showlegend = show_legend,
      hovertext = hover_txt, hoverinfo = "text"
    ) |>
    plotly::add_markers(
      x = t_peak, y = c_peak, name = "Median peak", showlegend = FALSE,
      marker = list(color = "#003f88", size = 10, symbol = "diamond"),
      hovertext = sprintf("Median peak<br>%.1f h · %.2f ng/mL", t_peak, c_peak),
      hoverinfo = "text"
    ) |>
    plotly::layout(
      title = list(text = ttl, font = list(size = 15), x = 0, xanchor = "left"),
      xaxis = list(title = "Time post-dose (hours)", automargin = TRUE),
      yaxis = y_axis,
      hovermode = "closest",
      legend = list(orientation = "h", x = 0, y = -0.38, yanchor = "top"),
      margin = list(t = 55, r = 20, b = 110, l = 80),
      font = list(family = story_font, size = 14),
      height = 560
    ) |>
    plotly::config(responsive = TRUE) |>
    htmlwidgets::onRender(fit_js)
}

build_fan(log_scale = FALSE, show_legend = TRUE)
```
:::

::: {.snap-panel}
## Population fan — log axis {#fan-log}

The same fan on a **log y-axis**. Concentrations span ~`r fmt(conc_span, 0)`× across the window, so the post-peak decline straightens into the near-linear elimination phase.

```{r}
build_fan(log_scale = TRUE, show_legend = TRUE)
```
:::

::: {.snap-panel}
## Variability by time point {#variability-by-time}

Spread and skew at each sampling time. The box shows the IQR; every jittered point is one subject.

```{r}
p_variability <- ggplot(pk_data, aes(x = factor(time_hr), y = concentration_ng_ml)) +
  geom_boxplot(fill = "#CFE8F3", color = "#4D4D4D", outlier.shape = NA, width = 0.58) +
  geom_jitter(width = 0.13, alpha = 0.25, size = 1, color = "#3A3A3A") +
  labs(
    x = "Time post-dose (hours)",
    y = "Concentration (ng/mL)"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_variability, tooltip = c("x", "y")) |>
  layout(
    margin = list(t = 55, r = 20, b = 65, l = 75),
    yaxis = list(automargin = TRUE),
    xaxis = list(automargin = TRUE),
    font = list(family = story_font, size = 14),
    height = 470
  ) |>
  config(responsive = TRUE) |>
  htmlwidgets::onRender(fit_js)
```
:::

::: {.snap-panel}
## Peak behavior — Cmax & Tmax {#peak-behavior}

Peak timing is remarkably consistent: **`r tmax_mode_pct`% of subjects peak at `r fmt(tmax_mode, 0)` h**, with the remainder at `r fmt(sort(unique(subject_peaks$tmax))[2], 0)` h. Because Tmax takes only a couple of discrete values here, a Cmax-vs-Tmax scatter with a trend line would be misleading — so we show the **Cmax distribution grouped by Tmax**, every subject a jittered point. Do the late-peaking subjects reach different concentrations?

```{r}
tmax_counts <- subject_peaks |>
  dplyr::count(tmax) |>
  dplyr::mutate(lab = paste0("Tmax = ", tmax, " h\n(n = ", n, ")"))
subject_peaks_lab <- subject_peaks |>
  dplyr::left_join(tmax_counts, by = "tmax")

p_peak <- ggplot(subject_peaks_lab,
                 aes(x = factor(tmax), y = cmax, text = paste0("Subject ", subject,
                     "\nTmax: ", tmax, " h\nCmax: ", sprintf("%.2f", cmax), " ng/mL"))) +
  geom_boxplot(aes(group = factor(tmax)), fill = "#CFE8F3", color = "#4D4D4D",
               outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.55, size = 1.6, color = "#D55E00") +
  labs(
    x = "Tmax — time of individual peak (hours)",
    y = "Cmax — individual peak concentration (ng/mL)"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_peak, tooltip = "text") |>
  layout(
    title = list(text = "Individual peak concentration by peak time (n = 50)",
                 font = list(size = 15), x = 0, xanchor = "left"),
    margin = list(t = 55, r = 20, b = 65, l = 75),
    yaxis = list(automargin = TRUE),
    xaxis = list(automargin = TRUE),
    font = list(family = story_font, size = 14),
    height = 470
  ) |>
  config(responsive = TRUE) |>
  htmlwidgets::onRender(fit_js)
```
:::

::: {.snap-panel}
## Normalized profile shape {#shape-comparison}

Normalizing each profile by its subject-level peak (`C/Cmax`) highlights shape differences independent of magnitude. Hover over any data point to see the subject.

```{r}
# Per-subject loop for normalized profiles — avoids ggplotly crash
p_norm <- plotly::plot_ly()
for (s in subjects) {
  d <- normalized_profiles[normalized_profiles$subject == s, ]
  ht <- paste0("<b>Subject ", s, "</b><br>Time: %{x} h<br>C/Cmax: %{y:.3f}<extra></extra>")
  p_norm <- plotly::add_trace(
    p_norm,
    x = d$time_hr, y = d$concentration_norm,
    name = s, type = "scatter", mode = "lines+markers",
    showlegend = FALSE,
    line = list(color = "rgba(80,80,80,0.35)", width = 1),
    marker = list(size = 10, opacity = 0),
    hovertemplate = ht
  )
}

# Add median overlay
p_norm <- p_norm |>
  plotly::add_lines(
    x = normalized_summary$time_hr, y = normalized_summary$median_norm,
    name = "Median", line = list(color = "#009E73", width = 2),
    showlegend = TRUE,
    hovertemplate = "Median C/Cmax at %{x} h: %{y:.3f}<extra></extra>"
  ) |>
  plotly::layout(
    title = list(text = "Normalized profile shape (C/Cmax) — hover to identify a subject (n = 50)",
                 font = list(size = 15), x = 0, xanchor = "left"),
    xaxis = list(title = "Time post-dose (hours)", automargin = TRUE),
    yaxis = list(title = "Normalized concentration (C/Cmax)", automargin = TRUE),
    hovermode = "closest",
    showlegend = TRUE,
    legend = list(orientation = "h", x = 0, y = -0.38, yanchor = "top"),
    margin = list(t = 55, r = 20, b = 110, l = 80),
    font = list(family = story_font, size = 14),
    height = 470
  )

p_norm |> plotly::config(responsive = TRUE) |>
  htmlwidgets::onRender(fit_js) |> htmlwidgets::onRender(highlight_js)
```
:::

::: {.snap-panel}
## Takeaways {#takeaways}

1. The median PK curve rises to a peak at **`r fmt(t_peak, 1)` h** (~`r fmt(c_peak, 1)` ng/mL) and declines over 24 hours, with an apparent terminal half-life of ~**`r fmt(thalf_med, 1)` h**.
2. Between-subject variability is visible at all time points, widest around the peak window; concentrations span ~`r fmt(conc_span, 0)`× across the study.
3. Peak timing is highly consistent (**`r tmax_mode_pct`% peak at `r fmt(tmax_mode, 0)` h**), and Cmax is relatively tight (median `r fmt(cmax_med, 1)` ng/mL [`r fmt(cmax_q25, 1)`–`r fmt(cmax_q75, 1)`]).
4. After normalization, profile shapes are broadly similar but still show meaningful heterogeneity, especially in the elimination phase.
:::
