---
title: 'Wonderful Wednesday Webinar #42'
subtitle: Dirty Data
date: "October 11, 2023"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  # This was updated to add a slope plot based on comments from the session
---


```{r, echo=FALSE, warning=FALSE, comment="#", error=FALSE, message=FALSE}





#        Program name: dirty_data.R 
#        Project name: WWW.2023.Oct.dirty_data 
#        Programmer name: Z. Skrivanek 
# 
#        Program creation date: Oct 5, 2023 
#        Program requirements: utils.R 
#        Called by:  NA
#        Output: NA 
#        Output location: NA 
#        Infrastructure: R 4.2.2, LINUX, hpc 
#        Validation Level: 1 
#        Comments:  this is for PSI Wonderful Wednesday Webinar.  This is all fake data. 
#           The data was generated in the SimClinicalTrials project under WWW.2023.September


#*******#
# SETUP #
#*******#

   if(.Platform$OS.type=="unix") {home <- system("cd ..;pwd",intern=T); tsep="/"}
   FUNCTIONS <- paste(home, "FUNCTIONS" ,sep=tsep)
  

 #****************#
 # ADMINISTRATIVE #
 # ASSIGNMENTS    #
 #****************#
  
   OUTPUT <- paste(home,"OUTPUT",sep=tsep)
   DATA <- paste(home, "DATA", sep=tsep)


 #****************#
 # ADMINISTRATIVE #
 # FUNCTIONS      #
 #****************#

   source(paste(FUNCTIONS, "utils.R",sep=tsep))
   library("tidyverse")

```
     
   Read in the data set.
   
```{r}
   
   PROdata <- read.csv(dfile("PROdata.csv"))
   head(PROdata)
   
```

Let's examine the demographic data to make sure there is nothing unusual

```{r}
  ggplot(PROdata) +
    geom_bar(aes(x=SEX, fill=TRT))
    

```

There are more males than females.  There do not seem to be any data errors, except for some missing values.

Now let's examine the ages of these patients for possible data errors.

```{r, warning=FALSE,error=FALSE, comment="#", message=FALSE}

  library(plotly)
 ## We want to match the colors in plotly with the colors in ggplot
  library(scales)    
  hex <- hue_pal()(2)
 
  plot_ly() %>%
    add_trace(data=PROdata,
              y = ~AGE,
              color = ~TRT,
              colors=hex,
              type="box",
              boxpoints="all",
              jitter=0.7,
              marker=list(size=5))
  

```
There do not appear to be any data errors in the age demographic.  Let's exmaine the Patient Reported Outcome questionnaire data.

We transform the data so that it is vertical regarding the questionnaires.  This will allow us to trellis the plots.

```{r}


  PROdata.vert <- PROdata %>%
    tidyr::pivot_longer(
      cols=starts_with("Q"),
      names_to = "Q_no",
      values_to = "result",
      names_prefix = "Q",
      values_drop_na=FALSE
    )

```

Let us compare the sites regarding the response to the questionnaires for the active and control groups.



```{r, fig.width=11, fig.align='center'}

  ggplot(PROdata.vert) +
     geom_histogram(aes(x = result, fill=TRT),
           binwidth=4,
           alpha=.5,
           position="identity") +
     labs(title = "Questionnaire Data",
           x = element_blank(),
           y= element_blank()) +
     facet_grid(SITE~Q_no) +
     theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.ticks.x=element_blank(), axis.ticks.y=element_blank()) +
     scale_y_continuous(breaks=NULL) + scale_x_continuous(breaks=NULL)
 

```

It appears that Site 4 is an outlier.  This is probably the site where the scores were reversed.  Based on comments from the October WWW session let's make a slope plot so we can really make it clear that this site is different than the other sites.

```{r, eval=FALSE}

## apply a plot to a data set where this works

  library(ggrepel)
# Loading data
df <- read_csv("https://raw.githubusercontent.com/GDS-ODSSS/unhcr-dataviz-platform/master/data/change_over_time/slope.csv")

# Plot
ggplot(df, aes(
  x = year,
  y = refugee_number,
  group = country_origin
)) +
  geom_line(
    size = 0.75,
    color = "blue" 
  ) +
  geom_text_repel(
    data = df |> filter(year == 2016),
    aes(label = paste(
      country_origin,
      if_else(refugee_number > 1e6,
        paste0(round(refugee_number / 1e6, 1), "M"),
        paste0(round(refugee_number / 1e3, 1), "k")
      )
    )),
    size = 8 / .pt,
    hjust = 1,
    direction = "y",
    nudge_x = -0.3
  ) +
  geom_text_repel(
    data = df |> filter(year == 2021),
    aes(label = paste(
      country_origin,
      if_else(refugee_number > 1e6,
        paste0(round(refugee_number / 1e6, 1), "M"),
        paste0(round(refugee_number / 1e3, 1), "k")
      )
    )),
    size = 8 / .pt,
    hjust = 0,
    direction = "y",
    nudge_x = 0.3
  ) +
  geom_point(
    size = 2.5,
    #color = unhcr_pal(n = 1, "pal_blue")
    color = "blue" 
  ) +
  labs(
    title = "Evolution of refugee population in East and\nHorn of Africa region | 2016 vs 2021",
    caption = "Source: UNHCR Refugee Data Finder\n© UNHCR, The UN Refugee Agency"
  ) +
  scale_x_continuous(
    breaks = c(2016, 2021),
    limits = c(2013, 2024)
  ) +
  scale_y_continuous(limits = c(-2e5, NA)) #+
  #theme_unhcr(
    #grid = "X",
    #axis = FALSE,
  #   axis_title = FALSE,
  #   axis_text = "X"
  #)




  
```

```{r}

## do the slope data viz for active to control by site

#*********#
# DATA    #
# MUNGING #
#*********#
 

  PROdata.slopes <- PROdata.vert %>%
    group_by(SITE, TRT) %>%
    summarise(total.score=sum(result, na.rm=TRUE))
  

#************#
# CREATE THE #
# DATA VIZ   #
#************#

## apply a plot to a data set where this works

  library(ggrepel)

p <- ggplot(PROdata.slopes, aes(
  x = TRT,
  y = total.score,
  group = SITE 
)) +
  geom_line(
    size = 0.75,
    color = "grey" 
  ) +
  geom_point(
    size = 2.5,
    #color = unhcr_pal(n = 1, "pal_blue")
    color = "grey" 
  ) +
  labs(
    title = "Total Scores By Site",
    caption = "Site 04 is Visual Outlier"
  ) +
  geom_text_repel(
    data = PROdata.slopes |> filter(TRT=="Active"),
    aes(label = paste(SITE, TRT,"=", total.score)),
    size = 8 / .pt,
    hjust = 1,
    direction = "y",
    nudge_x = -0.3,
  ) + 
  geom_text_repel(
    data = PROdata.slopes |> filter(TRT=="Control"),
    aes(label = paste(SITE, TRT,"=", total.score)),
    size = 8 / .pt,
    hjust = 1,
    direction = "y",
    nudge_x = 0.5,
  ) +
 ## Make Site 04 appear in blue
  geom_line(
    data=PROdata.slopes |> filter(SITE=="SITE04"),
    size = 0.75,
    color = "blue" 
  ) +
  geom_point(
    data=PROdata.slopes |> filter(SITE=="SITE04"),
    size = 2.5,
    #color = unhcr_pal(n = 1, "pal_blue")
    color = "blue" 
  ) + 
  xlab("Treatment") + ylab("Total Scores (Q1 + Q2 + Q3 + Q4)")

  ggsave(plot=p, ofile("slopes.png"))

```


Let's do some data munging to correct this data error.

```{r, fig.width=11}

   #========#
   # SWITCH #
   # SCORES #
   #========#
   
    flip.scale <- function(x){
     ## Simulate getting the scale wrong on a 5 point likert scale
     ## re-assign their values 1->5, 5->1, 2->4, 4->2, 3 <-> 3
      case_when( 
        x == 1 ~ 5,
        x == 5 ~ 1,
        x == 2 ~ 4,
        x == 4 ~ 2,
        x == 3 ~ 3
      )
    }

   ## apply the function
    site4.flipped.scores <- flip.scale(PROdata.vert$result[PROdata.vert$SITE=="SITE04"])
    
    tmp <- PROdata.vert
    tmp$result[tmp$SITE=="SITE04"] <- site4.flipped.scores
    
  ##  verify results
    
    ggplot(tmp) +
     geom_histogram(aes(x = result, fill=TRT),
           binwidth=4,
           alpha=.5,
           position="identity") +
     labs(title = "Questionnaire Data",
           x = element_blank(),
           y= element_blank()) +
     facet_grid(SITE~Q_no) +
     theme(axis.text.x=element_blank(), axis.text.y=element_blank(), axis.ticks.x=element_blank(), axis.ticks.y=element_blank()) +
     scale_y_continuous(breaks=NULL) + scale_x_continuous(breaks=NULL)
```

The data looks correct now. We will use this transformed data to create a plot of the scores, comparing the 2 treatments across all sites.


```{r, fig.width=11, warning=FALSE}

  PROdata.vert.corrected <- tmp
  write.csv(PROdata.vert.corrected, dfile("PROdata.vert.corrected.csv"))

 ## Summarize the data
  PROdata.sum <- PROdata.vert.corrected %>%
    group_by(TRT, Q_no) %>%
    count(result) %>%
    mutate(percent = n /sum(n))
  
 ## Save the data for use in RAWgraphics
  write.csv(PROdata.sum, dfile("PROdata.sum.csv"))
  
 ## Plot the data
  p <- ggplot(PROdata.sum, aes(x=result, y=percent, fill=TRT)) 
  p <- p + geom_bar(position='dodge',stat='identity') 
  
 ## label the facets
  Q.labs <- c("1" = "PRO Question 1",  "2" = "PRO Question 2", "3" = "PRO Question 3", "4" = "PRO Question 4"  )
  
  p <- p + facet_wrap(vars(Q_no), 
                 ncol=2,
                 labeller = labeller(Q_no=Q.labs))
  
 ## add title
  p <- p + ggtitle("Active group consistently scores higher than Control group\n on all PRO questions") +
    theme(plot.title=element_text(hjust=0.5))
  
  p + scale_y_continuous(labels = scales:: percent_format(accuracy = 1))

```


