---
title: "PSI Wonderfull Wednesday"
output: html_document
---

Thank you very much for the awesome event!
I'm sorry for the late submission and would greatly appreciate it if you could still consider it for a review.

## Libraries and data

```{r, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(ggplot2)
library(data.table)
library(readr)
library(viridis)

```

```{r}
# read data
urlfile <- "https://raw.githubusercontent.com/VIS-SIG/Wonderful-Wednesdays/master/data/2023/2023-09-13/PROdata.csv"
testData <- read_csv(url(urlfile))
setDT(testData)

# rename
names(testData) <- c("site", "id", "trt", "sex", "age", "q1", "q2", "q3", "q4")
```

## SITE04 swapped values

For each research site I created a histogram comparing answer distributions against all other sites. 
I observed that SITE04 had the most "opposite" answer distribution compared to the others. 
Considering this, I suspect that SITE04 may have swapped answer scales.

```{r}
# melt data
testData6 <- testData
names(testData6)[6:9] <- paste("Question", 1:4, sep = "")
testData4 <- melt(testData6, 
                  id.vars = c("site", "id", "trt", "sex", "age"), 
                  measure.vars = paste("Question", 1:4, sep = ""), 
                  variable.name = "question", 
                  value.name = "response")

# create plot
printPlot <- function(test, title){
  p1 <- ggplot(test,
               aes(x = response,
                   fill = site)) +
      geom_histogram(aes(y = ..density..), binwidth = 1, show.legend = F) +
      facet_grid(site ~ question) +
      scale_fill_viridis(discrete = T) +
      theme_minimal() +
      theme(axis.title = element_text(face = "bold"),
            plot.title = element_text(hjust = 0.5),
            strip.text.y = element_text(angle = 0))  +
      labs(title = "",
           y = "Density",
           x = "Response")
  
  print(p1)
}  

# split table
compareSites <- function(x){
  setDT(testData4)
  testData4With <- testData4[site == x,]
  testData4Without <- testData4[site != x,]
  testData4Without$site <- "Other Sites"
  testData5 <- rbind(testData4Without, testData4With)
  printPlot(testData5, x)
}

# # print all plots
sites <- unique(testData$site)
for(i in sites){
compareSites(i)
}

compareSites("SITE04")
  
```

## Exploring AGE

Unfortunately I couldn't identify any singular age values, which might have been unexpected. 
What surprised me, was the clear divide in age between men and women taking part in the study.

```{r}
testData2 <- na.omit(testData, cols = "sex")

# plot p3 histogram
p3 <- ggplot(testData2, aes(x = age,
                           fill = site)) +
  geom_histogram(binwidth = 1) +
  facet_grid(sex ~ .) +
  scale_fill_viridis(discrete = T,
                     label = paste("Site", 1:10, sep = "")) +
  theme_minimal() +
  theme(legend.position = "top",
        strip.text.y = element_text(angle = 0),
        axis.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey60", hjust = 0),
        plot.title = element_text(hjust = 0.5)) + 
  labs(fill = "",
       x = "Patient age",
       y = "Count",
       title = "",
       caption = "Excluding patients where information on sex is not available")

print(p3)
  
```

```{r}
testData2 <- na.omit(testData, cols = "sex")

# plot p3 histogram
p3 <- ggplot(testData2, aes(x = age,
                            fill = sex)) +
  geom_histogram(binwidth = 1) +
  facet_grid(sex ~ .) +
  scale_fill_viridis(discrete = T,
                     label = paste("Site", 1:10, sep = "")) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text.y = element_text(angle = 0),
        axis.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey60", hjust = 0),
        plot.title = element_text(hjust = 0.5)) + 
  labs(fill = "",
       x = "Patient age",
       y = "Count",
       title = "All males have age ≤ 50 and females age ≥ 50",
       caption = "Excluding patients where information on sex is not available")

print(p3)
  
```