---
title: "Identifying patient groups who might benefit from personalised dosing"
author: "Thomas Weissensteiner"
date: "2024-10-08"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 4
---
**My scripts:**  

* rendered html versions: [Rpubs/thomas-weissensteiner](https://rpubs.com/thomas-weissensteiner)  
* .rmd files with executable code chunks: www.github.com/thomas-weissensteiner/portfolio/tree/main/  

**About myself:** www.linkedin.com/in/ThomasWs-Mopfair
<br>

```{r Options, include = FALSE}
# - Global directory and output options                         - # 

# The working directory in "Global directory and output options" needs to be customised, and should include the following additional files:
# "variants_simulated.csv", the example data file
# "Summarizing_Human_Phenotype_Ontology_Terms_figs", a folder containing additional figures not generated by the code chunks  

knitr::opts_knit$set(root.dir = "C:/Users/ThomasWeissensteiner/OneDrive/Documents/portfolio/WWchallenge")
knitr::opts_chunk$set (warning = FALSE, message = F, tidy = FALSE)
knitr::opts_chunk$set (attr.output='style="max-height: 200px"')

```
<br>
<br>

### 1. Background and motivation
<br>

Current dosing practices are based on average responses from limited clinical trials. However, overlooking variations such as to sex, age, and genetics can lead to adverse events in certain patient groups. Conversely, a new drug which might have benefit for a sub-group of patients might fail to get approval because the pivotal trial showed insufficient general efficacy in the general population.

These scripts were written for the [VIS-SIG Wonderful Wednesdays challenge 10 Oct 24](https://github.com/VIS-SIG/Wonderful-Wednesdays/tree/master/data/2024/2024-09-11). [Wonderful Wednesdays](https://psiweb.org/sigs-special-interest-groups/visualisation/welcome-to-wonderful-wednesdays) are a monthly webinar organised by the [Visualisation Special Interest Group](https://psiweb.org/sigs-special-interest-groups/visualisation) of Statisticians in the Pharmaceutical Industry ([PSI](https://psiweb.org/)).  

The task consisted of identifying patient subgroups with dose responses that deviate from the remaining study participants, and who therefore might benefit from personalized dosing.
<br>
<br>

### 2. R packages and data set
<br>

```{r set-up}
# - set-up                                                     - #

library(dplyr)        # version 1.1.4  # general grammar
library(purrr)        # version 1.0.2  # set_names, list_rbind
library(tidyr)        # version 1.3.1  # pivot_longer, pivot_wider
library(ggpubr)       # version , includes ggplot2 3.4.0
                        # stat_compare_means, ggtexttable, ggarrange
library(kableExtra)   # rendering tables in HTML format 
                        # (kableExtra rather than kable is required for R 4.3.0)

exampleData <- read.csv("https://raw.githubusercontent.com/VIS-SIG/Wonderful-Wednesdays/refs/heads/master/data/2024/2024-09-11/WWW54ExampleData.csv")
```
<br>

The data were a simulated dose response set with the following structure:  

dose: dose levels (0, 100, 200 mg)  
target: continuous response variable (higher is better)  

subgroup variables:  
bmi (body mass index: ‘low BMI’ or ‘high BMI’)  
age (‘<40 years’ or ‘≥40 years’)  
race (‘Black’, ‘Asian’, ‘White’)  
sex (‘Female’, ‘Male’)  
type (type of disease: ‘Acute disease’ or ‘Chronic disease’)
<br>
<br>

### 3. Target vs. dose in all patients {#Fig1}
<br>

```{r chunk_1, fig.dim = c(7.5, 7.5)}
# - Chunk_1                                              - #
# - Requires R-libraries and object exampleData (set-up) - #

## Plot distribution of target values at different doses in the total population

sign_test <- 
  exampleData %>% 
  compare_means(
  data = .,
  target ~ dose,
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme(
      "light", 
      base_size = 10
      ),
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(
  
exampleData %>% 
  select (dose, target) %>% 
  
  ggplot (
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200"
        )
      ), 
      y = target)
    ) +
  geom_hline(
    yintercept = 0, 
    color = "darkgrey",
    linewidth = 1.1
    ) +
  geom_violin(
    color = "lightgrey",
    fill  = "lightgrey", 
    width = 0.45
    ) +
  geom_jitter(
    height = 0, width = 0.1, 
    size = 2, 
    alpha = 0.2
    ) +
  geom_boxplot(
    width = 0.15 , 
    outlier.shape = NA,
    color = "salmon",
    lwd = 0.7,
    alpha = 0.5
    ) +
  theme_light() +
  theme(
    text=element_text(size = 13), 
    axis.text = element_text(
      size = 12),
    ) +
  scale_y_continuous(
    breaks = c(-2: 5),
    minor_breaks = NULL
    ) +
  labs(
    x = "Dose", 
    y = "Target", 
    size = 12, hjust = -15,
    title = "Target vs. dose levels in the total patient population\n"
    ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1)
)   

```
<br>
<br>
The reference for the Wilcoxon test is the group of untreated patients (dose = 0). Differences in target levels in treated patients are significant. However, the distribution of target values at the three dose levels suggests that at least two subgroups might exists that could respond differently.
<br>
<br>

### 4. Best response in patient subgroups
<br>

In order to identify these potential groups, I stratified patients by one or two of the clinical and demographic characteristics that were provided.
For each group, I then calculated the median response for doses 100 and 200 (difference between median target values in the treated subgroups and patients receiving a dose 0).
<br>
<br>

#### 4.1. Table of responses in patient subgroups characterised by one or two variables  
<br>

```{r chunk_2}
# - Chunk_2                                              - #
# - Requires R-libraries and object exampleData (set-up) - #

# Generate all pairwise combinations of patient characteristics
pVars <- 
  exampleData %>% 
  names() %>% .[!. %in% c("dose","target")]

pVar_pairs <- 
  pVars %>% 
  { c(
    lapply(., function(x) c(x, x)),       # duplicate each characteristic
    combn(.,2, simplify = F)              # all combinations of 2 characteristics
  ) }
 
# ---
pChar_pairs <- 
  lapply(
    seq_along(pVar_pairs), 
    function (x) 
      exampleData %>% 
      select(pVar_pairs[[x]] ) %>% 
      unique(.)
    )
# ---

# Calculate median responses for the patient subgroups 
response_tbl <- 
  lapply(
    pVar_pairs, 
    function(x) 
      exampleData %>% 
        group_by(
          across(all_of(x) ), 
          dose
          ) %>% 
      summarise(
        median(target),
        N = n()
      )
  ) %>%
  # ensure all dataframes have the same number of columns 
    lapply(function(df)
      if(
        ncol(df) < length(pVars)) {
        cbind(df[, 1, drop = FALSE], df)    
        }else{
          df} 
      ) %>% 
  lapply(function(df)
    pivot_wider(
      df, 
      names_from = "dose",
      values_from = c("median(target)", "N")
      ) %>%
    set_names(
     "Var1_value", "Var2_value", 
     "median_0", "median_100", "median_200", 
      "N_0", "N_100", "N_200"
      )
    ) %>% 
  list_rbind() %>%
  mutate(
    response_100 = median_100 - median_0,
    response_200 = median_200 - median_0,
    max_response = max(response_100, response_200) ,
    max_resp_dose = which.max(c(response_100, response_200))*100,
    N_max = c(N_100, N_200)[max_resp_dose/100]
    )


## Table of best median responses stratified by patient subgroups

# Re-order columns, sort by maximum response (highest first)
response_tbl %>%   
  select( 
    "Var1_value", "Var2_value", 
    "N_max", "max_response",
    "N_0", "N_100", "N_200",
    "median_0", "median_100", "median_200"
    ) %>% 
  arrange( desc(max_response)) %>% 
  kable(., "html") %>% 
  scroll_box(width = "100%", height = "200px") %>% 
  kable_styling( 
    bootstrap_options = c(
      "striped", "hover", "responsive", full_width = F) 
    ) 


```
<br>
<br>

#### 4.2. Best responses in patients stratified by demographic and clinical parameters - showing patient numbers in treated and matched untreated subgroups {#Fig2}  
<br>

```{r chunk_3, fig.dim = c(15,15)}
# - Chunk_3                                              - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object response_tbl (chunk_2)               - #

# Optional: rearrange so that diagonal of the plot shows responses in order of magnitude
response_tbl <- 
  response_tbl %>%
  mutate(
    .before = Var1,
    Var1 = case_when (
      Var1_value %in% c("high BMI", "low BMI") ~ "BMI",
      Var1_value %in% c(">=40 years", "<40 years") ~ "Age",
      Var1_value %in% c("Asian", "Black", "White") ~ "Ethnic",
      Var1_value %in% c("Chronic disease", "Acute disease") ~ "Type",
      Var1_value %in% c("Female", "Male") ~ "Sex"
      ),
    Var2 = case_when (
      Var2_value %in% c("high BMI", "low BMI") ~ "BMI",
      Var2_value %in% c(">=40 years", "<40 years") ~ "Age",
      Var2_value %in% c("Asian", "Black", "White") ~ "Ethnic",
      Var2_value %in% c("Chronic disease", "Acute disease") ~ "Type",
      Var2_value %in% c("Female", "Male") ~ "Sex"
      )
    ) %>% 
    mutate(
      pVar_equal = (Var1 == Var2)
      ) %>%  
    { rbind( 
        filter(., pVar_equal) %>% 
          group_by(Var1) %>%
          arrange(desc(max_response), .by_group = TRUE) %>%
          mutate(max_value = max(max_response, na.rm = TRUE)) %>%
          arrange(desc(max_value)) %>% 
          select (!max_value),
        filter(., !pVar_equal) 
        ) 
      } 

# Select and reshape  

response_tbl %>% 
  select(
    Var1,
    Var1_value, Var2_value, 
    N_max, N_0, 
    max_resp_dose, max_response
    ) %>% 
  pivot_wider(
    names_from = "Var2_value",
    values_from = c(
      "max_resp_dose", "max_response", "N_max", "N_0"), 
      names_sep = "-"
      ) %>% 
  pivot_longer(
    cols = !c("Var1", "Var1_value"),
    names_to = c("resp", "Var2_value"),
    values_to = c("max_response"), names_sep = "-"
    ) %>% 
  pivot_wider(
    names_from = "resp", 
    values_from = "max_response") %>% 
  mutate(
    Var1 = as.factor(Var1)
    ) %>% 

# Plot data
  ggplot(
    aes(
      x = Var1_value %>% 
        factor(., levels = unique(.) %>% rev ), 
      y = Var2_value %>% 
        factor(., levels = unique(.) %>% rev ),
      col   = max_response,
      label = max_response %>% round(3) 
      )
    ) +
    geom_tile(
      aes(fill= factor(max_resp_dose)),
      col="lightgrey"
      ) +
    scale_fill_manual(
      values = c("100" = "#BBE0CF", "200" = "#D5F9E8"),
      na.translate = F
      ) +
    geom_text(
      col = "black", 
      vjust = 2.5, hjust = 0.5, 
      size = 10/.pt
      ) +
    geom_point(
      aes( size = N_0), 
      shape = 15, 
      color = "white",
      position = position_nudge(y = 0.1)
      ) +
    geom_point(
      aes( size = N_max), 
      shape = 19,
      position = position_nudge(y = 0.1)
      ) +
    geom_point(
      aes( size = N_0), 
      shape = 22, 
      color = "darkblue",
      position = position_nudge(y = 0.1)
      ) +
    scale_size(range = c(1, 15), name = "N") +
    labs(
      x = NULL, y = NULL, 
      col = "Maximum\nresponse",
      fill = "Dose",
      title = "Best median response, stratified by one or two patient characteristics") +
    scale_color_gradient2(
      low = "white", high="darkblue", 
      limits = c(-0.25, 1.5)
      ) +
    scale_x_discrete(
      position = "top"
      ) +
    scale_y_discrete() +
    theme_light(base_size = 15) + 
    theme(
      axis.text.x = element_text(
        angle = 90, 
        vjust = 0.5,
        hjust = 0
        ),
      panel.grid = element_blank() 
    )

```
<br>

Tiles in the plot correspond to patient groups characterised by the row and column labels.
Magnitude of the maximum median response in each patient group is shown by the colour of the dots and the numerical values below them. The numbers (N) of patients in the different groups is indicated by the size of the dots (treated), and white squares with dark blue border (untreated).
The dose that achieved the maximum median response is shown by the colour of the tiles (dark green: 100, lighter green: 200).  
<br>

The diagonal shows groups stratified by a single characteristic, arranged so that the magnitude of the responses increases from bottom left to top right. The acute disease type was associated with the highest response, followed by female sex, and White ethnic. Conversely, patients with chronic disease, or were male or Black, had the lowest responses. BMI and age had relatively smaller effects as a single stratifier.  

When two characteristics were applied, females <40 years had the highest response, followed by Whites with acute disease, White females, and females or Blacks with acute disease.
The group wit the lowest median response were patients with chronic disease and high BMI. Low responses were also observed in patients who were Black and male, <40 years, with chronic disease, or high BMI. This is in contrast to the Blacks with acute disease having among the highest responses. However, numbers in these groups were relatively small and the result might need confirmation in a larger sample.

In most groups, the higher dose (200) was associated with the greater response, but Asians with chronic disease or high BMI, or patients with high BMI and acute disease might benefit more from the lower dose.
<br>
<br>

#### 4.3. Best response in patients stratified by demographic and clinical parameters - showing Wilcoxon / Mann-Whitney test p-values for difference between treated and matched untreated subroups {#Fig3} 
<br>

```{r chunk_4a}
# - Chunk_4a                                                  - #
# - Requires R-libraries and object exampleData (set-up)      - #
# - Requires objects response_tbl and pChar_pairs (chunk_2)   - #


## Calculate Wilcoxon test p-values for all comparisons

wilcox_p_value <- function(x, y) {
  test <- wilcox.test(x, y)
  return(test$p.value)
}

wilcox.test_results <- list ()

for (i in 1: length(pChar_pairs)) {
  wilcox.test_results_j <- list()

  for (j in 1: nrow(pChar_pairs[[i]])) {
    x <- pChar_pairs[[i]] %>% names
    y <- pChar_pairs[[i]] %>% .[j, ]%>% set_names(x)
    # Ensure names are preserved when cbinding as transformed dataframe below
    if (length(x) == 1) {
      y <- as.data.frame (t(y))
      }

  wilcox.test_results_j[[j]] <- 
    exampleData %>% 
      {if (length(x) == 1) {
        filter(.,
          get(x[1]) == paste(y[1])  
          ) 
        } else {
          filter(.,
            get(x[1]) == paste(y[1]) & 
            get(x[2]) == paste(y[2])
            ) 
          } 
        } %>% 
      summarise(
        p_value_0_vs_100 = wilcox_p_value(target[dose == "0"], target[dose == "100"] ),
        p_value_0_vs_200 = wilcox_p_value(target[dose == "0"], target[dose == "200"] )
        ) %>% 
      cbind(y, .)
    }
    wilcox.test_results[[i]] <- wilcox.test_results_j

  }

wilcox.test_results <- 
wilcox.test_results %>% 
  flatten() %>% 
  lapply(function(df)
      if(
        ncol(df) < 4) {
        cbind(df[, 1, drop = FALSE], df)    
        }else{
          df} 
      ) %>% 
    lapply(function(df)
      set_names(df,
        "Var1_value", "Var2_value", 
        "p_value_0_vs_100", "p_value_0_vs_200"
        ) 
      ) %>% 
    list_rbind() %>% 
    mutate(
      log.p_100 = log10(p_value_0_vs_100) %>% round(digits = 2),
      log.p_200 = log10(p_value_0_vs_200) %>% round(digits = 2)
    ) %>% 
    select(
      !c(p_value_0_vs_100, p_value_0_vs_200)
    )

```


```{r chunk_4b, fig.dim = c(15,15)}
# - Chunk_4b                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object wilcox.test_results (chunk_4a)       - #

wilcox.test_results  <- 
  wilcox.test_results %>% 
    unite(
      Vars1_2,
      c(Var1_value, Var2_value), 
      sep = "_", 
      remove = F
      )

response_tbl %>% 
  unite(
    Vars1_2,
    c(Var1_value, Var2_value), 
    sep = "_", 
    remove = F
    ) %>% 
  left_join(
    wilcox.test_results) %>% 
  mutate(
     log.p_max.resp = case_when(
      max_resp_dose == "100" ~ log.p_100, 
      max_resp_dose == "200" ~ log.p_200
     )
    ) %>% 
  
# Select and reshape  
  select(
    #Var1,
    Var1_value, Var2_value, 
    log.p_max.resp, 
    max_resp_dose, max_response
    ) %>% 
  pivot_wider(
    names_from = "Var2_value",
    values_from = c(
      "max_resp_dose", "max_response", "log.p_max.resp"), 
      names_sep = "-"
      ) %>% 
  pivot_longer(
    cols = !c("Var1", "Var1_value"),
    names_to = c("resp", "Var2_value"),
    values_to = c("max_response"), names_sep = "-"
    ) %>% 
  pivot_wider(
    names_from = "resp", 
    values_from = "max_response") %>% 
  mutate(
    Var1_value = as.factor(Var1_value)
    ) %>% 

# Plot data
  ggplot(
    aes(
      x = Var1_value %>% 
        factor(., 
          levels = unique(.) %>% rev
          ), 
      y = Var2_value %>% 
        factor(., 
          levels = unique(.) %>% rev 
          ),
      col   = max_response,
      label = max_response %>% round(3) 
      )
    ) +
    geom_tile(
      aes(fill= factor(max_resp_dose)),
      col="lightgrey"
      ) +
    scale_fill_manual(
      values = c("100" = "#BBE0CF", "200" = "#D5F9E8"),
      na.translate = F
      ) +
    geom_text(
      col = "black", 
      vjust = 3.5, hjust = 0.5, 
      size = 10/.pt
      ) +
    geom_point(
      aes( size = log.p_max.resp), 
      shape = 19,
      position = position_nudge(y = 0.1)
      ) +
    scale_size(
      range = c(18, 1), 
      name = "log.p_max.resp") +
    labs(
      x = NULL, y = NULL, 
      col = "Maximum\nresponse",
      fill = "Dose",
      title = "Best median response, stratified by one or two patient characteristics",
      subtitle = "\nlog.p: Wilcox test"
      ) +
    scale_color_gradient2(
      low = "white", high="darkblue", 
      limits = c(-0.25, 1.5)
      ) +
    scale_x_discrete(
      position = "top"
      ) +
    scale_y_discrete() +
    theme_light(base_size = 15) + 
    theme(
      axis.text.x = element_text(
        angle = 90, 
        vjust = 0.5,
        hjust = 0
        ),
      panel.grid = element_blank() 
      )

wilcox.test_results %>% 
  select(!Vars1_2) %>% 
  kable(., "html") %>% 
    scroll_box(width = "50%", height = "150px") %>% 
    kable_styling( 
      font_size = 10,
      bootstrap_options = "striped"
      )

```
<br>
<br>

#### 4.4. Best response in patients stratified by demographic and clinical parameters - showing Mood test p-values for difference between treated and matched untreated subroups {#Fig4}   
<br>

```{r chunk_5a}
# - Chunk_5a                                                  - #
# - Requires R-libraries and object exampleData (set-up)      - #
# - Requires objects response_tbl and pChar_pairs (chunk_2)   - #

## Calculate Mood test p-values for all comparisons

mood_p_value <- function(x, y) {
  test <- mood.test(x, y)
  return(test$p.value)
}

mood.test_results <- list ()

for (i in 1: length(pChar_pairs)) {
  mood.test_results_j <- list()

  for (j in 1: nrow(pChar_pairs[[i]])) {
    x <- pChar_pairs[[i]] %>% names
    y <- pChar_pairs[[i]] %>% .[j, ]%>% set_names(x)
    # Ensure names are preserved when cbinding as transformed dataframe below
    if (length(x) == 1) {
      y <- as.data.frame (t(y))
      }

  mood.test_results_j[[j]] <- 
    exampleData %>% 
      {if (length(x) == 1) {
        filter(.,
          get(x[1]) == paste(y[1])  
          ) 
        } else {
          filter(.,
            get(x[1]) == paste(y[1]) & 
            get(x[2]) == paste(y[2])
            ) 
          } 
        } %>% 
      summarise(
        p_value_0_vs_100 = mood_p_value(target[dose == "0"], target[dose == "100"] ),
        p_value_0_vs_200 = mood_p_value(target[dose == "0"], target[dose == "200"] )
        ) %>% 
      cbind(y, .)
    }
    mood.test_results[[i]] <- mood.test_results_j

  }

mood.test_results <- 
mood.test_results %>% 
  flatten() %>% 
  lapply(function(df)
      if(
        ncol(df) < 4) {
        cbind(df[, 1, drop = FALSE], df)    
        }else{
          df} 
      ) %>% 
    lapply(function(df)
      set_names(df,
        "Var1_value", "Var2_value", 
        "p_value_0_vs_100", "p_value_0_vs_200"
        ) 
      ) %>% 
    list_rbind() %>% 
    mutate(
      log.p_100 = log10(p_value_0_vs_100) %>% round(digits = 2),
      log.p_200 = log10(p_value_0_vs_200) %>% round(digits = 2)
    ) %>% 
    select(
      !c(p_value_0_vs_100, p_value_0_vs_200)
    )

```


```{r chunk_5b, fig.dim = c(15,15)}
# - Chunk_5b                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object wilcox.test_results (chunk_5a)       - #

mood.test_results  <- 
  mood.test_results %>% 
    unite(
      Vars1_2,
      c(Var1_value, Var2_value), 
      sep = "_", 
      remove = F
      )

response_tbl %>% 
  unite(
    Vars1_2,
    c(Var1_value, Var2_value), 
    sep = "_", 
    remove = F
    ) %>% 
  left_join(
    mood.test_results) %>% 
  mutate(
     log.p_max.resp = case_when(
      max_resp_dose == "100" ~ log.p_100, 
      max_resp_dose == "200" ~ log.p_200
     )
    ) %>% 
  
# Select and reshape  
  select(
    #Var1,
    Var1_value, Var2_value, 
    log.p_max.resp, 
    max_resp_dose, max_response
    ) %>% 
  pivot_wider(
    names_from = "Var2_value",
    values_from = c(
      "max_resp_dose", "max_response", "log.p_max.resp"), 
      names_sep = "-"
      ) %>% 
  pivot_longer(
    cols = !c("Var1", "Var1_value"),
    names_to = c("resp", "Var2_value"),
    values_to = c("max_response"), names_sep = "-"
    ) %>% 
  pivot_wider(
    names_from = "resp", 
    values_from = "max_response") %>% 
  mutate(
    Var1_value = as.factor(Var1_value)
    ) %>% 

# Plot data
  ggplot(
    aes(
      x = Var1_value %>% 
        factor(., 
          levels = unique(.) %>% rev
          ), 
      y = Var2_value %>% 
        factor(., 
          levels = unique(.) %>% rev 
          ),
      col   = max_response,
      label = max_response %>% round(3) 
      )
    ) +
    geom_tile(
      aes(fill= factor(max_resp_dose)),
      col="lightgrey"
      ) +
    scale_fill_manual(
      values = c("100" = "#BBE0CF", "200" = "#D5F9E8"),
      na.translate = F
      ) +
    geom_text(
      col = "black", 
      vjust = 3.5, hjust = 0.5, 
      size = 10/.pt
      ) +
    geom_point(
      aes( size = log.p_max.resp), 
      shape = 19,
      position = position_nudge(y = 0.1)
      ) +
    scale_size(
      range = c(18, 1), 
      name = "log.p_max.resp") +
    labs(
      x = NULL, y = NULL, 
      col = "Maximum\nresponse",
      fill = "Dose",
      title = "Best median response, stratified by one or two patient characteristics",
      subtitle = "\nlog.p: Mood test"
      ) +
    scale_color_gradient2(
      low = "white", high="darkblue", 
      limits = c(-0.25, 1.5)
      ) +
    scale_x_discrete(
      position = "top"
      ) +
    scale_y_discrete() +
    theme_light(base_size = 15) + 
    theme(
      axis.text.x = element_text(
        angle = 90, 
        vjust = 0.5,
        hjust = 0
        ),
      panel.grid = element_blank() 
      )

mood.test_results %>% 
  select(!Vars1_2) %>% 
  kable(., "html") %>% 
    scroll_box(width = "50%", height = "150px") %>% 
    kable_styling( 
      font_size = 10,
      bootstrap_options = "striped"
      )

```
<br>
<br>

### 5. Distribution of target values in patient subgroups treated with different doses  
<br>

```{r chunk_6a}
# - Chunk_6a                                                                          - #
# - From: https://stackoverflow.com/questions/35717353/split-violin-plot-with-ggplot2 - #

## Custom ggproto object for generating split violin plots

GeomSplitViolin <- ggplot2::ggproto(
    "GeomSplitViolin",
    ggplot2::GeomViolin,
    draw_group = function(self,
                          data,
                          ...,
                          # add the nudge here
                          nudge = 0,
                          draw_quantiles = NULL) {
        data <- transform(data,
                          xminv = x - violinwidth * (x - xmin),
                          xmaxv = x + violinwidth * (xmax - x))
        grp <- data[1, "group"]
        newdata <- plyr::arrange(transform(data,
                                           x = if (grp %% 2 == 1) xminv else xmaxv),
                                 if (grp %% 2 == 1) y else -y)
        newdata <- rbind(newdata[1, ],
                         newdata,
                         newdata[nrow(newdata), ],
                         newdata[1, ])
        newdata[c(1, nrow(newdata)-1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

        # now nudge them apart
        newdata$x <- ifelse(newdata$group %% 2 == 1,
                            newdata$x - nudge,
                            newdata$x + nudge)

        if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {

            stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <= 1))

            quantiles <- ggplot2:::create_quantile_segment_frame(data,
                                                             draw_quantiles)
            aesthetics <- data[rep(1, nrow(quantiles)),
                               setdiff(names(data), c("x", "y")),
                               drop = FALSE]
            aesthetics$alpha <- rep(1, nrow(quantiles))
            both <- cbind(quantiles, aesthetics)
            quantile_grob <- ggplot2::GeomPath$draw_panel(both, ...)
            ggplot2:::ggname("geom_split_violin",
                             grid::grobTree(ggplot2::GeomPolygon$draw_panel(newdata, ...),
                                            quantile_grob))
        }
    else {
            ggplot2:::ggname("geom_split_violin",
                             ggplot2::GeomPolygon$draw_panel(newdata, ...))
        }
    }
)

geom_split_violin <- function(mapping = NULL,
                              data = NULL,
                              stat = "ydensity",
                              position = "identity",
                              # nudge param here
                              nudge = 0,
                              ...,
                              draw_quantiles = NULL,
                              trim = TRUE,
                              scale = "area",
                              na.rm = FALSE,
                              show.legend = NA,
                              inherit.aes = TRUE) {

    ggplot2::layer(data = data,
                   mapping = mapping,
                   stat = stat,
                   geom = GeomSplitViolin,
                   position = position,
                   show.legend = show.legend,
                   inherit.aes = inherit.aes,
                   params = list(trim = trim,
                                 scale = scale,
                                 # don't forget the nudge
                                 nudge = nudge,
                                 draw_quantiles = draw_quantiles,
                                 na.rm = na.rm,
                                 ...))
}

```


I used split violin box plots for the pairwise comparison of target medians and distributions in low and high responder groups. Low responders are shown in green on the left side, high responder groups in red on the right.
Likewise, facets for the second group variable are arranged from lowest (left) to highest responders (right).

<br>
<br>

#### 5.1. Dose vs. target values in patients stratified by sex and age {#Fig5}  
<br>
(see also 3.6.: female patients stratified by sex and age)

```{r chunk_6b, fig.dim = c(10, 7.5)}
# - Chunk_6b                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# - Requires R-libraries and objects exampleData (chunk: set-up) and GeomSplitViolin (chunk 5) - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- compare_means(
  data = exampleData,
  target ~ dose,
  group.by = c("sex", "age"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("sex", "age", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Sex", "Age", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select( dose, target) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200") 
        ), 
      y = target,
# Order low > high responders
    fill = 
      factor(
        sex, 
        levels = c("Male", "Female") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue( direction = -1) +    
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
# Order facets low > high responders
    facet_grid(. 
      ~ factor(
        age, 
        levels = c(">=40 years", "<40 years")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Sex",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 2),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

#### 5.2. Dose vs. target values in patients stratified by sex and type of disease {#Fig6}    

```{r chunk_6c, fig.dim = c(10, 7.5)}
# - Chunk_6c                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- compare_means(
  data = exampleData,
  target ~ dose,
  group.by = c("sex", "type"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("sex", "type", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Sex", "Type", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select( dose, target) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200") 
        ), 
      y = target,
# Order low > high responders
    fill = 
      factor(
        sex, 
        levels = c("Male", "Female") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue( direction = -1) +    
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
 # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        type, 
        levels = c("Chronic disease", "Acute disease")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Sex",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1.5),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

#### 5.3. Dose vs. target values in patients stratified by ethnic and type of disease {#Fig7} 
<br>
(see also 3.5.: female patients stratified by ethnic and type of disease)

```{r chunk_6d, fig.dim = c(10, 7.5)}
# - Chunk_6d                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #


# Generate a table of significant differences (Wilcoxon test) 

sign_test <- compare_means(
  data = exampleData,
  target ~ dose,
  group.by = c("type", "race"),
  ref.group = "0",
 ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select(!".y." & !"group1" & !"method")  %>% 
  set_names(
    "Sex", "Ethnic", "Dose", "p", "p.adj", "p.format", "p.signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by( bmi, age, race, sex, type) %>% 
  select( dose, target) %>% 
  ggplot(
    aes(
      x= factor(dose, levels = c("0", "100", "200") ), 
      y = target,
# Order low > high responders
    fill = factor (type, levels = c("Chronic disease", "Acute disease") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue (direction = -1) +
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
 # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        race, 
        levels = c("Black", "Asian", "White")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Type",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 3),
widths = c(0.2, 1, 1),
align = "h"
)    
  
```
<br>
<br>
Interestingly, Blacks with chronic disease had high target levels, even when untreated,
<br>
<br>

#### 5.4. Dose vs. target values in patients stratified by race and sex {#Fig8}    
<br>

```{r chunk_6e, fig.dim = c(10, 7.5)}
# - Chunk_6e                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- compare_means(
  data = exampleData,
  target ~ dose,
  group.by = c("sex", "race"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select(!".y." & !"group1" & !"method")  %>% 
  set_names(
    "Sex", "Ethnic", "Dose", "p", "p.adj", "p.format", "p.signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select (dose, target) %>% 
  ggplot(
    aes(
      x= factor(dose, levels = c("0", "100", "200") ), 
      y = target,
# Order low > high responders
    fill = 
      factor(
        sex, 
        levels = c("Male", "Female") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue( direction = -1) +    
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
  # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        race, 
        levels = c("Black", "Asian", "White")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Sex",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 2),
widths = c(0.2, 1, 1),
align = "h"
)    
  
```
<br>
<br>

#### 5.5. Dose vs. target values in female patients stratified by ethnic and disease type {#Fig9}  
<br>

```{r chunk_6f, fig.dim = c(10, 7.5)}
# - Chunk_6f                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- 
  exampleData %>% 
  filter(sex == "Female") %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = c("race", "type"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("race", "type", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Ethnic", "Type", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select (dose, target) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200") 
        ), 
         y = target,
# Order low > high responders
    fill = factor (type, levels = c("Chronic disease", "Acute disease") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue (direction = -1) +       geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
 # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        race, 
        levels = c("Black", "Asian", "White")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Type",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 3),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

#### 5.6. Dose vs. target values in female patients stratified by ethnic and age {#Fig10}  
<br>

```{r chunk_6g, fig.dim = c(10, 7.5)}
# - Chunk_6g                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- 
  exampleData %>% 
  filter(sex == "Female") %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = c("race", "age"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("race", "age", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Ethnic", "Age", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select (dose, target) %>% 
  ggplot (
    aes (x= factor(dose, levels = c("0", "100", "200") ), 
         y = target,
# Order low > high responders
        fill = factor (age, levels = c(">=40 years", "<40 years") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue (direction = -1) +
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
 # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        race, 
        levels = c("Black", "Asian", "White")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target",
      fill = "Age",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 2.5),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

Younger white females appear to respond better than older, especially at the higher dose, whereas the opposite appears to be true for Asian females (not sure about the significance of that...).
<br>
<br>

#### 5.7. Dose vs. target values in White patients stratified by sex and disease type {#Fig11}  
<br>

```{r chunk_6h, fig.dim = c(10, 7.5)}
# - Chunk_6h                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- 
  exampleData %>% 
  filter(race == "White") %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = c("sex", "type"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("sex", "type", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Sex", "Type", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select (dose, target) %>% 
  ggplot (
    aes (x= factor(dose, levels = c("0", "100", "200") ), 
         y = target,
# Order low > high responders
    fill = 
      factor(
        sex, 
        levels = c("Male", "Female") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue( direction = -1) +    
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
 # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        type, 
        levels = c("Chronic disease", "Acute disease")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Sex",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1.5),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

#### 5.8. Dose vs. target values in White patients stratified by sex and age {#Fig12}  
<br>

```{r chunk_6i, fig.dim = c(10, 7.5)}
# - Chunk_6i                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- 
  exampleData %>% 
  filter(race == "White") %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = c("sex", "age"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("sex", "age", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Sex", "Age", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select (dose, target) %>% 
  ggplot(
    aes(
      x = factor(
        dose, 
        levels = c("0", "100", "200") 
        ), 
      y = target,
# Order low > high responders
    fill = 
      factor(
        sex, 
        levels = c("Male", "Female") )
        ) 
      ) +
# Color red for patient variable value with higher response 
    scale_fill_hue( direction = -1) +    
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
     geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
# Order facets low > high responders
    facet_grid(. 
      ~ factor(
        age, 
        levels = c(">=40 years", "<40 years")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "Sex",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1.5),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

#### 5.9. Dose vs. target values in patients stratified by race and BMI {#Fig13} 
<br>
(The difference between median targets in the Black/ low BMI/ dose 100 group had the highest significance in the Mood test)  

```{r chunk_6k, fig.dim = c(10, 7.5)}
# - Chunk_6k                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

sign_test <- 
  exampleData %>% 
  filter(race == "White") %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = c("race", "bmi"),
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("race", "bmi", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Ethnic", "BMI", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme("light")
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 
    

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  group_by(bmi, age, race, sex, type) %>% 
  select(dose, target) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200")
        ), 
      y = target,
      fill = bmi
      ) 
    ) +
# Color red for patient variable value with higher response 
    scale_fill_hue (direction = -1) +
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
     geom_split_violin( 
      width = 1,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
        width = 0.3, 
        outlier.shape = NA) +
  # Order facets low > high responders
    facet_grid(. 
      ~ factor(
        race, 
        levels = c("Black", "Asian", "White")
        ) 
      ) +
    theme_bw() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) + 
    labs(
      x = "Dose", 
      y = "Target", 
      fill = "BMI",
      size = 12, 
      hjust = -15
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

### 6. Groups with the highest and lowest responses
<br>

It appears that the groups the received the most significant benefit were women who were either <40 years (p.adj = 7.9 * 10^-13^, median difference 1.474), White (p.adj = 1.2 * 10^-12^, median difference 1.406), or with acute disease (p.adj = 3.9 * 10^-12^, median difference 1.407). 
Whites with acute disease or <40 years also showed significant benefit (p.adj = 2.7 * 10^-10^, median difference 1.421; median difference 1.474, 3.3 * 10^-9^, median difference 1.377), but this could be mainly because of the female sub-population of these groups.  
<br>

#### 6.1. White females <40 years respond better than the remaining patients {#Fig14}
<br>

```{r chunk_7a, fig.dim = c(10, 7.5)}
# - Chunk_7a                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

exampleData <- 
  exampleData %>% 
    mutate(
      Group = 
        case_when(
          race == "White" &
          #type == "Acute disease" &
          age == "<40 years" &
          sex == "Female"
             ~ "White / F / <40 yrs",
        TRUE ~ "Remaining"
        )
      )

sign_test <- 
  exampleData %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = "Group",
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("Group", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Group", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme(
      "light", 
      base_size = 10
      ),
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  select (dose, target, type, Group) %>% 
  group_by(Group) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200")
        ), 
      y = target,
      fill = Group 
      ) 
    ) +
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    fill_palette(
      palette = c("lightgrey","salmon")
      ) +
    geom_split_violin( 
      width = 1,
      nudge = 0.01,
      alpha = 0.3,
      color = "grey") +
    geom_boxplot(
      width = 0.2, 
      outlier.shape = NA) +
    geom_point(
      pch = 21,
      size = 3,
      position = 
        position_jitterdodge(
          dodge.width = 0.7, 
          jitter.width = 0.1
          ),
      color = "grey",
      alpha = 0.5
      ) +
    theme_light() +
    theme(
      text=element_text( size = 13), 
      axis.text = element_text( size = 12),
      ) +
    scale_y_continuous(
      breaks = c(-2: 5),
      minor_breaks = NULL
      ) +    
    labs(
      x = "Dose", 
      y = "Target", 
      size = 13, hjust = -15,
      title = "Target vs. dose levels in the group with highest responses\n"
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1),
widths = c(0.2, 1, 1),
align = "h"
)    

```
             
<br>
<br>

#### 6.2. Black males with chronic disease show no or negative response {#Fig15}
<br>
However, this might need confirmation in a larger sample of this subgroup.
<br>

```{r chunk_7b, fig.dim = c(10, 7.5)}
# - Chunk_7b                                             - #
# - Requires R-libraries and object exampleData (set-up) - #
# - Requires object geom_split_violin (chunk_6a)         - #

# Generate a table of significant differences (Wilcoxon test) 

exampleData <- 
  exampleData %>% 
    mutate(
      Group = 
        case_when(
          race == "Black" &
          # bmi == "high BMI" &
           type == "Chronic disease" &
          # age == ">=40 years" &
          sex == "Male"
             ~ "Black/ M / Chronic disease",
        TRUE ~ "Remaining"
        )
      )

sign_test <- 
  exampleData %>% 
  compare_means(
  data = .,
  target ~ dose,
  group.by = "Group",
  ref.group = "0",
  ) %>%
  filter (
    !group1 == "100" &
    !p.signif == "ns"  ) %>% 
  select("Group", "group2", "p.format", "p.adj", "p.signif")  %>% 
  set_names(
    "Group", "Dose", "p", "p.adj", "signif") %>% 
  ggtexttable(
    rows = NULL,
    theme = ttheme(
      "light", 
      base_size = 10
      ),
    ) %>% 
  tab_add_title(
    text = "Wilcoxon test (ref. dose = 0)", 
    face = "bold", size = 12,
    ) 

# Generate split violin plot and insert the Wilcoxon test table 

ggarrange(

exampleData %>% 
  select (dose, target, type, Group) %>% 
  group_by(Group) %>% 
  ggplot(
    aes(
      x= factor(
        dose, 
        levels = c("0", "100", "200")
        ), 
      y = target,
      fill = Group 
      ) 
    ) +
    geom_hline(
      yintercept = 0, 
      color = "darkgrey",
      linewidth = 1.1
      ) +
    fill_palette(
      palette = c("turquoise", "lightgrey")
      ) +
    geom_split_violin( 
      width = 1,
      nudge = 0.01,
      alpha = 0.3,
      color = "grey") +
  geom_boxplot(
      width = 0.2, 
      outlier.shape = NA) +
  geom_point(
    pch = 21,
    size = 3,
    position = 
      position_jitterdodge(
        dodge.width = 0.7, 
        jitter.width = 0.1
        ),
    color = "grey",
    alpha = 0.5
    ) +
  theme_light() +
  theme(
    text=element_text( size = 13), 
    axis.text = element_text( size = 12),
    ) +
  scale_y_continuous(
    breaks = c(-2: 5),
    minor_breaks = NULL
    ) +    labs(
      x = "Dose", 
      y = "Target", 
      size = 13, hjust = -15,
      title = "Target vs. dose levels in the group with lowest responses\n"
      ),

# Add space between plot and table
NULL,

sign_test,

ncol = 1, nrow = 3,
heights = c(2, 0.1, 1),
widths = c(0.2, 1, 1),
align = "h"
)    

```
<br>
<br>

### 7. Question from a non-statistician
<br>
The [distribution of target in the total population](#Fig1) looks similar for all 3 doses, and suggests that at least two major subgroups exist that might respond differently.  

In the following 3 figures I tried to produce an overview of [best responses in subgroups characterised by 1-2 of the patient variables](#Fig2). Because there was no guarantee for a linear relationship between dose and target, I decided to compare the dose 100 and dose 200 groups individually with the dose 0 (untreated) group.
Because in many subgroups the target values did not seem to be normally distributed, I choose to compare the differences between the median target values. 
I also tried to assess the significance of these differences. It struck me that the that the p-values of the [Wilcoxon test](#Fig3) seemed extremely significant in all groups, even after correction. By contrast, the [Mood test](#Fig4) produced no significant results.  

So how can we find out which of responses are significant? A partial answer might be in the next figures, which again show [distributions of target at the 3 doses, but for individual subgroups](#Fig5). As I understand, the Wilcoxon (but not the Mood) test requires the the distributions of the compared groups are "similar". Visually inspecting plots for each of the subgroups characterised by 2 variables would seem subjective and cumbersome. I am out of my depth here, and grateful for learning what could be the most useful approach.  
